Canopy leaf area of an evergreen Eucalyptus woodland does not respond to elevated

### 2015-05-04 manuscript.Rmd
More refs; bug fixes; author info moved to top

Leaf area index (*L*) is an important ecosystem attribute underlying productivity of vegetation, [@Reich2012], and a key variable in land surface models [@Kala2013; @Walker2014]. *L* is a dynamic canopy property, as it is generally responsive to variation in water and nutrient availability, as demonstrated in experiments [@Herbert1995; @Hubbard2004; @White2010] and observational studies across rainfall gradients [@Ellis2008; @Iio2013; @Jin2013]. Ecosystem models generally predict that *L* increases in response to elevated atmospheric [CO~2~] (eC~a~) due to increased net carbon uptake [@DeKauwe2014; @Hatton1992; @Kergoat2002; @Zhu2012]. Also expected because increased quantum yield of photosynthesis in eC~a~ would allow leaf retention into deeper shade [@Norby1999], but little empirical evidence is available on the response of *L* to eC~a~ in mature forests [@Norby2011]. Quantifying the magnitude and direction of the eC~a~ effect on *L* is crucial, as changes in *L* associated with climate change drivers such as increases in C~a~ may provide an important feedback to climate change, and modify direct physiological responses to C~a~ [@Tor-ngern2015; @Murray2012].

\
Direct effects of eC~a~ on vegetation processes include an enhancement of photosynthesis and productivity, a decrease in stomatal conductance and leaf-level water use and an increase in the water-use efficiency [@Ainsworth2005; @Ellsworth2012]. Although numerous studies have been conducted on the response of physiology and plant growth to eC~a~, very few of these inform the response of steady-state *L* to eC~a~, because it requires a sufficiently mature forest to be exposed to free air CO~2~ enrichment (FACE) [@Norby2011]. Increased carbon availability through enhanced photosynthesis under eC~a~ does not necessarily lead to increases in *L*. Out of the six forest FACE studies that have been carried out worldwide, only broad-leaved deciduous aspen stands (in monoculture or mixed with birch) and needle-leaved evergreen pine stands showed increased *L* in response to eC~a~ *L* [@Norby2011], although positive responses to eC~a~ in terms of productivity were observed at all six sites. Interestingly, Norby and Zak [-@Norby2011] found a negative correlation between pre-treatment *L* and the responsiveness to eC~a~, leading to the prediction that stands with low *L* will show a larger response to eC~a~. Because diminished returns in increasing *L* in response to eC~a~ when LAI is larger [@Norby2005].

\
In ecosystems where water availability limits productivity, we could expect eC~a~ to further affect *L* [@Donohue2009], since eC~a~ reduces stomatal conductance, and hence water use per unit of leaf area [@Warren2010] leading to increased soil water availability. These ‘water savings’ can be expected to lead to an increase in *L* in water-limited ecosystems under the idea that leaf area is closely coupled to water-balance. **Explain here or just before that we expect this because of the tight relationship between LAI and rainfall across gradients.** In support of this prediction, Donohue et al. [-@Donohue2013] attributed the observed global increase in *L* in arid and semi-arid areas over the last 30 years to rising C~a~. Based on the same water-savings principle, we predict that *L* should also increase under eC~a~ in water-limited forests.

\
Most previous studies on *L* responses to environmental drivers have focused on the seasonal, mostly temperature and light-driven, timing of *L* dynamics observed in N-hemisphere forests (e.g. [@Richardson2012; @Savoy2015]) whereas *L* responses to fluctuations in water availability are qualitatively well-known (**more cites**), but poorly quantified. The *L* of broad-leaved evergreen trees that dominate forests in the southern hemisphere show particular sensitivity to water-availability [@Pook1985; @Specht1989; @Whitehead2004; @Pook1997; @Limousin2009].The strong relationship between *L* and rainfall observed across Australia [@Ellis2008; @Specht1989], and other continents [@Donohue2013; @Grier1977; @Jin2013] represents the adaptation of vegetation structure and density to long-term climate conditions. However, local *L* is also highly responsive to water availability at shorter time scales [@Pook1985; @Pook1997], driven by leaf turnover [@Misson2011] and leaf flushing and expansion [@Muller2011], which are both highly responsive to soil water availability. Nevertheless, few datasets are currently available at sufficiently high temporal resolution to allow modelling of water-driven changes in phenology [@Misson2011].

\
We estimated *L* from near-continuous measurements of canopy transmittance in a Eucalyptus-dominated woodland exposed to eC~a~, the EucFACE experiment. The woodland stand at the EucFACE is frequently water-limited, allowing us to quantify the dynamics of *L*, and its components (litter fall and production) in response to water availability. We hypothesized that eC~a~ would result in an increase in *L*, particularly during periods of water limitation, as water savings were expected to lead to longer retention of foliage under elevated CO~2~. In addition, we expected that the low *L* that is typical of native *Eucalyptus*-dominated woodlands [@Whitehead2004] allowed a larger response to eC~a~ [@Norby2011]. Our second objective was to quantify the temporal dynamics of *L*, in particular in relation to fluctuating water availability.

@@

## *L* from diffuse transmittance
In each of the six FACE rings, 3 photosynthetically active radiation (PAR) sensors (LI-190, Li-cor, Lincoln, NE, U.S.A.) were installed on metal posts at 1 m height similar to the approach used by Norby et al. [-@Norby2003]. Data were logged on data loggers (CR3000, Campbell Scientific Australia, Townsville, Australia) at minutely intervals from 15 Oct. 2012, and automatically uploaded every day to a data management system. At each ring, another PAR sensor was installed above the canopy, on a mast extending above a scaffolding tower in the centre of each ring at 23.5 m height. All PAR data were averaged in 30 min time steps. At a nearby site (within 2 km), we measured the diffuse fraction (Fdiff) of PAR at 30 min time step using a BF5 Sunshine sensor (Delta-T Instruments, Cambridge, UK) installed on a tower extending above the canopy. For determination of *L*, we used only highly diffuse conditions, where Fdiff > 0.98. We further trimmed the data by excluding low solar elevation (< 10°), because, for unknown reasons, measured canopy transmittance was more variable at these low solar angles. Finally canopy transmittance was calculated for every 30-min time step as,

(2) $$\tau_d = \bar{Q}_{below} / Q_{above}$$

@@

(3) $$\tau_d = 2\int_0^{\pi/2} e^{-LG(\theta)/cos(\theta)}sin(\theta)cos(\theta)d\theta$$

where $\theta$ is the solar zenith angle, G($\theta$) is the foliage projection function in the direction of the beam [-@Campbell2000]. Because we assumed a spherical leaf angle distribution, G($\theta$) = 0.5 [@Stenberg2006]. We solved numerically for *L* in Eq. 3, using a smooth integration over $\theta$, yielding $L_\tau$.

\
To overcome violations in the assumptions underlying Eq. 3, we calibrated *L* against litter production. We estimated *L* from $\tau_d$ ($L_\tau$) and calibrating against cumulative litter production during a drought period in 2013 (indicated in Fig. 1). We selected a period of sustained drought in 2013 (14 July 2013 – 12 Nov 2013), during which no leaf growth occurred but *L* decreased steadily (and thus $\tau_d$ increased steadily, see Fig. 1). During this period, only 42 mm precipitation was recorded, which was the second driest July-October period recorded at a nearby (<2km) weather station in operation since 1881 (compared to a long-term average of 188 mm for this period). A total of n=11 days with sufficiently diffuse conditions to estimate $\tau_d$  were available during this period. In theory, cumulative *L* leaf litter production ($P_L$, expressed in units of *L*) during that period should thus match the observed decrease in *L* estimated from canopy transmittance. We calibrated Eq. 4,

(4) $$c(L_{\tau,t} - L_{\tau,t=0}) = \int_0^t P_Ldt$$

where *c* is a calibration parameter, $L_{\tau,t=0}$ is $L_\tau$ at the start of the drought period. We used linear regression without intercept to estimate *c* in Eq. 4 as `r calib` (SE = `r tidy(caliblm)$std.error`). It should be stressed that even after calibration with litter production, $L_\tau$ still includes stem and infrastructure area (see Terminology), because this approximately constant area cancels out in the left-hand side of Eq. 4.

@@

We took advantage of the detailed timeseries of *L* estimates from canopy transmittance to test a simple photography-based method proposed by Macfarlane et al. [-@Macfarlane2007], which is extensively applied to *Eucalyptus*-dominated forests. From October 2012 to October 2013, we took ca. monthly canopy photographs in each FACE ring. Photographs were taken in the morning on cloudy days. We took one photograph at each point of a 4 x 4 m grid established within each ring, unless obstructed by trees, yielding 20-23 photos per ring. Sampling points too close to either the vertical vent pipes (edges of the ring) or the central scaffolding tower were excluded. The camera was aligned with the north-south axis and set at 1.25 m height.  Camera lens was pointed directly upwards and levelled using a three axis bubble level. We used a Nikon D90 camera with a Nikon AF 50 mm lens (Nikon Corporation, Tokyo, Japan). To measure cover, the camera was set to automatic exposure, Aperture-Priority and automatic autofocus mode and F10 lens. Following recommendations by Jonckheere et al. [-@Jonckheere2005], we used the automated Ridler algorithm to threshold the canopy photos. We then calculated canopy transmittance ($\tau_p$) from each canopy photo as the proportion of pixels with intensity in the blue band below that threshold. *L* from canopy photos ($L_p$) was calculated with a simple Beer-Lambert law (Eq. 1).

(1) $$L_p = -log(\tau_p)/k$$

where $\tau_p$ canopy transmission estimated from canopy photos, and we assumed that the extinction coefficient k = 0.5, following Macfarlane et al. [-@Macfarlane2007] recommendation for eucalypt forests.

\
We compared $L_\tau$ (after calibration to litter production) to the independent estimate of *L* from canopy photos ($L_p$). Without further calibration, the two estimates were reasonably well correlated, though on average $L_\tau$ was `r 100*(with(flatcan_byring, mean(LAI.PAR.mean / LAI))-1)` higher than $L_p$ (Fig. SI2, $R^2$=`r glance(lmphoto)$r.squared`, RMSE=`r glance(lmphoto)$sigma`). This is encouraging because the photos measure *L* near the zenith (+/- 15°), while $L_\tau$ is integrated over the sky hemisphere (Eq. 3), weighted towards the zenith.

@@

# Results

Canopy transmittance ($\tau_d$) varied between `r min(df$Gapfraction.mean)` and `r max(df$Gapfraction.mean)`, with much temporal variation, and substantial absolute differences between the individual FACE rings (Fig. 1).  However, temporal trends were very similar between the rings. Since the understorey comprises mostly grasses and forbs, and the PAR sensors were installed at 1 m height, this variation is due to the overstorey trees only.

@@

\
The dynamics of $L_\tau$ were compared with litter fall, by estimating the absolute change in $L_\tau$ for every FACE ring, for every period when litter was collected. We did this by fitting a generalized additive model to $L_\tau$ for each ring separately (similar to shown for $\tau_d$ in Fig. 1), and calculating the change in $L_\tau$ for each period ($\Delta L_\tau$). Litter production correlated significantly with $\Delta L_\tau$ when the latter was separated into periods where $\Delta L_\tau$ > 0 (net leaf growth) and $\Delta L_\tau$ <  0 (net leaf loss) (Fig. 4). During periods of net leaf loss, we found a significant correlation (with a slope of `r tidy(lmlitdown)$estimate[2]`, SE = `r tidy(lmlitdown)$estimate[2]`, n = `glance(lmlitdown)$df.residual+2`), close to the 1:1 line, as expected since no leaf growth was observed during any period where *L* was decreasing. During periods of net leaf growth, we found a significant correlation (with a slope of `r tidy(lmlitup)$estimate[2]`, SE = `r tidy(lmlitup)$estimate[2]`, n = `glance(lmlitup)$df.residual+2`) between $\Delta L_\tau$ and litter production (Fig. 5).

@@

\
Finally, we quantified the relationship between timing of leaf flushing and environmental drivers. During the `r nummon` month study period, we observed four leaf flushing events in the field, with one before estimates of $L_\tau$ were available (Fig. 7). These events occurred in spring and summer, with no clear annual cycle (Oct. 2012, Jan. 2013, Dec. 2013 and Jan. 2015). The dynamics of $L_\tau$ from the sensor record matched these events observed in the field quite well (Fig. 7a), as estimated from a generalized additive model fitted to the average $L_\tau$ across the six FACE rings. From the fitted model, we estimated the numerical derivative of *L* (dL/dt), as well as its confidence interval. All four leaf flushing events corresponded to significant positive dL/dt. One brief significant positive dL/dt (June 2013) did not correspond to an actual leaf flushing event, but it was small in magnitude. It is also obvious that every leaf flush event occurred shortly after a heavy rainfall event (Fig. 4 and 6), when the soil water content in the top 80 cm increased to ca. > 0.2 $m^3$ $m^{-3}$. Two periods of sustained decrease in *L* (red lines in Fig. 7b, during winter of 2013 and 2014) corresponded to periods of low soil water availability.

@@

## Response of *L* to elevated [CO~2~]
We report on `r nummon` months of continuous leaf area index (*L*) measurements in a native evergreen *Eucalyptus* woodland exposed to +150ppm elevated CO~2~ (eC~a~) and 3 months of ramp-up in C~a~. We had predicted that *L* would increase in response to eC~a~ as a result of increased quantum yield for photosynthesis, enhanced CO~2~ assimilation for the entire canopy, and increased soil water availability. We also expected that this eucalypt woodland, with its inherently low *L* [@Whitehead2004], would be even more responsive to eC~a~ than higher-*L* and more closed-canopy stands [@Norby2011]. Counter to our expectations but similar to Norby et al. [-@Norby2003], we found no significant effect of elevated C~a~ on *L*. It could be argued that the change in *L* in response to eC~a~ was too small to be detected. Natural variation in *L* among plots in a native woodland is indeed a challenge to manipulative studies with low replication, such as FACE experiments. To increase sensitivity and statistical robustness, we used a novel method to account for pre-treatment differences by using stand basal area as a covariate. Using this method, we estimated that we can detect a difference of only a `r format(Ldetect,digits=1)`% in *L*. Based on this small detectable difference, we conclude that the lack of a significant C~a~ effect is not due to a lack of power, but a true lack of response to elevated C~a~.

@@

\
Other studies in FACE are consistent with longer leaf-lifespan in elevated Ca shown here.

@@

**Figure 1**. (a) Target [CO~2~] (C~a~) relative to the ambient C~a~ treatment at the EucFACE. The elevated C~a~ treatment proceeded in steps of 30 ppm, reaching +150 ppm in Feb. 2013. (b) Canopy transmittance ($\tau_d$) estimated from continuous time series of photosynthetically active radition (PAR) below and above the canopy, for highly diffuse conditions ($F_diff$ > 0.98) only. The solid lines are smoothed time series (with 95% confidence intervals) for each FACE ring (using a generalized additive model). Lines in blue colours denote ambient C~a~ treatment, red colours are elevated C~a~ (+150 ppm). Note steep increase in $\tau_d$ for Ring 6 around November 2014, due to an insect attack that resulted in near defoliation of the canopy. The arrow labelled ‘calibration’ indicates a drought period when $\tau_d$ steadily increased, used to calibrate *L* estimates to litter production (see Methods).

**Figure 2**. Average $L_\tau$ over the entire study period plotted against the basal area (sum of stem cross-sectional areas) of the FACE rings ($R^2$ = `r glance(balm)$r.squared`, *L* = `tidy(balm)$estimate[1]` + `tidy(balm)$estimate[2]` BA). This relationship was used to calculate the *L* anomaly in Fig. 3, as the difference between *L* predicted from this regression and the actual $L_\tau$ at any given time.

@@

**Figure 6**.  (a) Total litter production (converted from data in Fig. 5 to an annual estimate) for each of the FACE rings plotted against mean $L_\tau$ during the entire litter collection period. The ratio of mean $L_\tau$ to annual litter production gives the mean leaf life span. Grey lines indicate leaf lifespan isolines. (b) Calculated leaf lifespan for the six FACE rings shows a significant difference between C~a~ treatments (p = `r pval(LLtest$p.value)`).

**Figure 7**. Dynamics of *L* in relation to environmental drivers. (a) Smoothed time series of *L_\tau* averaged across all six FACE rings. (b) From panel (a), we calculated the daily change in *L* (dL/dt), including the 95% confidence interval for the derivative of the time series. Green indicates significant positive dL/dt, red indicates significant negative dL/dt (at $\alpha$ = 0.05). Four of the five significant positive increases corresponded to approximate dates of commencement of leaf flushing (horizontal lines) observed in the field (except the third brief significant increase). (c) Soil water content estimated from over 0-80 cm soil depth interval, averaged across all sensors and all rings (blue line), and daily rainfall (black lines).  (d) Average daily air temperature with a fitted smoother.

@@

**Figure S2**. Correlation between *L* estimated from diffuse canopy transmittance ($L_\tau$) and from canopy photos (L~p~). Solid line is a 1:1 line, dashed line a regression line across all data (with the shading indicating the 95% confidence interval), and the colored lines regressions per plot across dates. (a) Estimates of L~p~ with the extinction coefficient (*k*) set to 0.5, as recommended by MacFarlane et al. [-@Macfarlane2007] ($L_\tau$ = `r tidy(lmcanphot)$estimate[1]` + `r tidy(lmcanphot)$estimate[2]` L~p~, R^2^ = `r glance(lmcanphot)$r.squared`, p `r pval(glance(lmcanphot)$p.value)`), (b) Estimates of L~p~ with *k* optimized to minimize the difference between L~p~ and $L_\tau$ ($L_\tau$ = `r tidy(lmcanphot)$estimate[1]` + `r tidy(lmcanphot)$estimate[2]`).

**Figure S3.** Comparison of canopy transmittance ($\tau$) during highly diffuse conditions ($F_diff$ > 0.98), or using all PAR data without filtering. Shown are daily average $\tau$ for all six FACE rings (symbols), with a smoother (generalized additive model with 95% confidence intervals). Note the much lower $\tau$ during southern hemisphere winter when using all data compared to diffuse only. This difference arises due to low solar angles, which lead to higher attenuation of the direct beam as a result of a longer path length through the canopy.

>>>>

Leaf area index (*L*) is an important ecosystem attribute underlying productivity of vegetation [@Reich2012], and a key variable in land surface models [@Kala2013; @Walker2014]. *L* is a dynamic canopy property, as it is generally responsive to variation in water and nutrient availability, as demonstrated in experiments [@Herbert1995; @Hubbard2004; @White2010] and observational studies across rainfall gradients [@Ellis2008; @Iio2013; @Jin2013].  Ecosystem models generally predict that *L* increases in response to  due to increased net carbon uptake [@DeKauwe2014; @Hatton1992; @Kergoat2002; @Zhu2012]. Other mechanisms are likely involved as well, including increased quantum yield of photosynthesis in eC~a~ allowing leaf retention into deeper shade [@Norby1999], and interactions with improved water relations under eC~a~  [@Wullschleger2002]. Quantifying the magnitude and direction of the effect of elevated atmospheric [CO~2~] (eC~a~) on *L* is crucial, as changes in *L* associated with climate change drivers such as increases in C~a~ may provide an important feedback to climate change, and modify direct physiological responses to C~a~ [@Tor-ngern2015; @Murray2012]. To date, little empirical evidence is available on the response of *L* to eC~a~ in mature forests [@Norby2011].

\
Direct effects of eC~a~ on vegetation processes include an enhancement of photosynthesis and productivity, a decrease in stomatal conductance and leaf-level water use and an increase in the water-use efficiency [@Ainsworth2005; @Ellsworth2012]. Although numerous studies have been conducted on the response of physiology and plant growth to eC~a~, very few of these inform the response of steady-state *L* to eC~a~, because it requires a sufficiently mature forest to be exposed to free air CO~2~ enrichment (FACE) [@Norby2011]. Increased carbon availability through enhanced photosynthesis under eC~a~ does not necessarily lead to increases in *L*. Out of the six forest FACE studies that have been carried out worldwide, only broad-leaved deciduous aspen stands (in monoculture or mixed with birch) and needle-leaved evergreen pine stands showed increased *L* in response to eC~a~ *L* [@Norby2011], although positive responses to eC~a~ in terms of productivity were observed at all six sites. Interestingly, Norby and Zak [-@Norby2011] found a negative correlation between pre-treatment *L* and the responsiveness to eC~a~, leading to the prediction that stands with low *L* will show a larger response to eC~a~.

\
In ecosystems where water availability limits productivity, we could expect eC~a~ to further affect *L* [@Donohue2009], since eC~a~ reduces stomatal conductance, and hence water use per unit of leaf area [@Warren2010] leading to increased soil water availability. These ‘water savings’ can be expected to lead to an increase in *L* in water-limited ecosystems, since we observe a tight coupling of *L* with precipitation across gradients [@Ma2013; @Jin2013; @Hatton2008]. In support of this prediction, Donohue et al. [-@Donohue2013] attributed the observed global increase in *L* in arid and semi-arid areas over the last 30 years to rising C~a~. Based on the same water-savings principle, we predict that *L* should increase under eC~a~ in water-limited forests.

\
Most previous studies on *L* responses to environmental drivers have focused on the seasonal, mostly temperature and light-driven, timing of *L* dynamics observed in N-hemisphere forests [e.g. @Richardson2012; @Savoy2015] whereas *L* responses to fluctuations in water availability are qualitatively well-known [@Pook1985; @Specht1989; @Whitehead2004; @Pook1997; @Limousin2009], but poorly quantified. The strong relationship between *L* and rainfall observed across Australia [@Ellis2008; @Specht1989; @Broich2014], and other continents [@Donohue2013; @Grier1977; @Jin2013] represents the adaptation of vegetation structure and density to long-term climate conditions. However, local *L* is also highly responsive to water availability at shorter time scales [@Pook1985; @Pook1997], driven by leaf turnover [@Misson2011] and leaf flushing and expansion [@Muller2011], which are both highly responsive to soil water availability. Nevertheless, few datasets are currently available at sufficiently high temporal resolution to allow modelling of water-driven changes in phenology [@Misson2011].

\
We estimated *L* from near-continuous measurements of canopy transmittance in a *Eucalyptus*-dominated woodland exposed to eC~a~, the EucFACE experiment. The woodland at the EucFACE is frequently water-limited, allowing us to quantify the dynamics of *L*, and its components (litter fall and production) in response to water availability. We hypothesized that eC~a~ would result in an increase in *L*, particularly during periods of water limitation, as water savings were expected to lead to longer retention of foliage under elevated CO~2~. In addition, we expected that the low *L* that is typical of native *Eucalyptus*-dominated woodlands [@Whitehead2004] allowed a larger response to eC~a~ [@Norby2011]. Our second objective was to quantify the temporal dynamics of *L*, in particular in relation to fluctuating water availability.

@@

## *L* from diffuse transmittance
In each of the six FACE rings, 3 photosynthetically active radiation (PAR) sensors (LI-190, Li-cor, Lincoln, NE, U.S.A.) were installed on metal posts at 1 m height similar to the approach used by Norby et al. [-@Norby2003]. Data were logged on data loggers (CR3000, Campbell Scientific Australia, Townsville, Australia) at minutely intervals from 15 Oct. 2012, and automatically uploaded every day to a data management system. At each ring, another PAR sensor was installed above the canopy, on a mast extending above a scaffolding tower in the centre of each ring at 23.5 m height. All PAR data were averaged in 30 min time steps. At a nearby site (within 2 km), we measured the diffuse fraction (F~diff~) of PAR at 30 min time step using a BF5 Sunshine sensor (Delta-T Instruments, Cambridge, UK) installed on a tower extending above the canopy. For determination of *L*, we used only highly diffuse conditions, where F~diff~ > 0.98. We further trimmed the data by excluding low solar elevation (< 10°), because, for unknown reasons, measured canopy transmittance was more variable at these low solar angles. Finally canopy transmittance was calculated for every 30-min time step as,

\
(1) 
$$\tau_d = \bar{Q}_{below} / Q_{above}$$

@@

\
(2)
$$\tau_d = 2\int_0^{\pi/2} e^{-LG(\theta)/cos(\theta)}sin(\theta)cos(\theta)d\theta$$

where $\theta$ is the solar zenith angle, G($\theta$) is the foliage projection function in the direction of the beam [-@Campbell2000]. Because we assumed a spherical leaf angle distribution, G($\theta$) = 0.5 [@Stenberg2006]. We solved numerically for *L* in Eq. 2, using a smooth integration over $\theta$, yielding $L_\tau$.

\
To overcome violations in the assumptions underlying Eq. 2, we calibrated *L_\tau* against cumulative litter production during a drought period in 2013 (indicated in Fig. 1). During this period (14 July 2013 – 12 Nov 2013), no leaf growth occurred but *L* decreased steadily (and thus $\tau_d$ increased steadily, see Fig. 1). During the drought, only 42 mm precipitation was recorded, which was the second driest July-October period recorded at a nearby (<2km) weather station since 1881 (compared to a long-term average of 188 mm). A total of n=11 days with sufficiently diffuse conditions to estimate $\tau_d$  were available during this period. Cumulative *L* leaf litter production ($P_L$, expressed in units of *L*) should match the observed decrease in *L* estimated from canopy transmittance. We calibrated Eq. 3,

\
(3) 
$$c(L_{\tau,t} - L_{\tau,t=0}) = \int_0^t P_Ldt$$

where *c* is a calibration parameter, $L_{\tau,t=0}$ is $L_\tau$ at the start of the drought period. We used linear regression without intercept to estimate *c* in Eq. 4 as `r calib` (SE = `r tidy(caliblm)$std.error`). It should be stressed that even after calibration with litter production, $L_\tau$ still includes stem and infrastructure area (see Terminology), because this approximately constant area cancels out in the left-hand side of Eq. 3. We used a single calibration constant for all rings, because we did not find significant differences between the rings in terms of either litter production or change in $\tau_d$ during the 2013 drought (see Fig. S1).

@@

We took advantage of the detailed timeseries of *L* estimates from canopy transmittance to test a simple photography-based method proposed by Macfarlane et al. [-@Macfarlane2007], which is extensively applied to *Eucalyptus*-dominated forests. From October 2012 to October 2013, we took ca. monthly canopy photographs in each FACE ring. Photographs were taken in the morning on cloudy days. We took one photograph at each point of a 4 x 4 m grid established within each ring, unless obstructed by trees, yielding 20-23 photos per ring. Sampling points too close to either the vertical vent pipes (edges of the ring) or the central scaffolding tower were excluded. The camera was aligned with the north-south axis and set at 1.25 m height.  Camera lens was pointed directly upwards and levelled using a three axis bubble level. We used a Nikon D90 camera with a Nikon AF 50 mm lens (Nikon Corporation, Tokyo, Japan). To measure cover, the camera was set to automatic exposure, Aperture-Priority and automatic autofocus mode and F10 lens. Following recommendations by Jonckheere et al. [-@Jonckheere2005], we used the automated Ridler algorithm to threshold the canopy photos. We then calculated canopy transmittance ($\tau_p$) from each canopy photo as the proportion of pixels with intensity in the blue band below that threshold. *L* from canopy photos ($L_p$) was calculated with a simple Beer-Lambert law (Eq. 4).

\
(4) 
$$L_p = -log(\tau_p)/k$$

where $\tau_p$ canopy transmission estimated from canopy photos. We assumed that the extinction coefficient *k* = 0.5, following Macfarlane et al. [-@Macfarlane2007] recommendation for eucalypt forests. Without further calibration, the two estimates were reasonably well correlated, though on average $L_\tau$ was `r 100*(with(flatcan_byring, mean(LAI.PAR.mean / LAI))-1)` higher than $L_p$ (Fig. SI2, $R^2$=`r glance(lmphoto)$r.squared`, RMSE=`r glance(lmphoto)$sigma`).

, and compared this assumption with an optimized extinction that minimized the difference between the two estimates of *L*.

\
We compared $L_\tau$ (after calibration to litter production) to the independent estimate of *L* from canopy photos ($L_p$).  This is encouraging because the photos measure *L* near the zenith (+/- 15°), while $L_\tau$ is integrated over the sky hemisphere (Eq. 3), weighted towards the zenith.

@@

# Results
Canopy transmittance ($\tau_d$) varied between `r min(df$Gapfraction.mean)` and `r max(df$Gapfraction.mean)`, with much temporal variation, and substantial absolute differences between the individual FACE rings (Fig. 1).  However, temporal trends were very similar between the rings. Since the understorey comprises mostly grasses and forbs, and the PAR sensors were installed at 1 m height, this variation is due to the overstorey trees only.

@@

\
The dynamics of $L_\tau$ were compared with litter fall, by estimating the absolute change in $L_\tau$ for every FACE ring, for every period when litter was collected. We did this by fitting a generalized additive model to $L_\tau$ for each ring separately (similar to shown for $\tau_d$ in Fig. 1), and calculating the change in $L_\tau$ for each period ($\Delta L_\tau$). Litter production correlated significantly with $\Delta L_\tau$ when the latter was separated into periods where $\Delta L_\tau$ > 0 (net leaf growth) and $\Delta L_\tau$ <  0 (net leaf loss) (Fig. 4). During periods of net leaf loss, we found a significant correlation (with a slope of `r tidy(lmlitdown)$estimate[2]`, SE = `r tidy(lmlitdown)$estimate[2]`, n = `r glance(lmlitdown)$df.residual+2`), close to the 1:1 line, as expected since no leaf growth was observed during any period where *L* was decreasing. During periods of net leaf growth, we found a significant correlation (with a slope of `r tidy(lmlitup)$estimate[2]`, SE = `r tidy(lmlitup)$estimate[2]`, n = `r glance(lmlitup)$df.residual+2`) between $\Delta L_\tau$ and litter production (Fig. 5).

@@

\
Finally, we quantified the relationship between timing of leaf flushing and environmental drivers. During the `r nummon` month study period, we observed four leaf flushing events in the field, with one before estimates of $L_\tau$ were available (Fig. 7). These events occurred in spring and summer, with no clear annual cycle (Oct. 2012, Jan. 2013, Dec. 2013 and Jan. 2015). The dynamics of $L_\tau$ from the sensor record matched these events observed in the field quite well (Fig. 7a), as estimated from a generalized additive model fitted to the average $L_\tau$ across the six FACE rings. From the fitted model, we estimated the numerical derivative of *L* (dL/dt), as well as its confidence interval. All four leaf flushing events corresponded to significant positive dL/dt. One brief significant positive dL/dt (June 2013) did not correspond to an actual leaf flushing event, but it was small in magnitude. It is also obvious that every leaf flush event occurred shortly after a heavy rainfall event (Fig. 4 and 6), when the soil water content in the top 80 cm increased to ca. > 0.2 m^3^ m^-3^. Two periods of sustained decrease in *L* (red lines in Fig. 7b, during winter of 2013 and 2014) corresponded to periods of low soil water availability.

@@

## Response of *L* to elevated [CO~2~]

** tie to objectives at end of intro! **

We report on `r nummon` months of continuous leaf area index (*L*) measurements in a native evergreen *Eucalyptus* woodland exposed to +150ppm elevated CO~2~ (eC~a~) and 3 months of ramp-up in C~a~. We had predicted that *L* would increase in response to eC~a~ as a result of increased quantum yield for photosynthesis, enhanced CO~2~ assimilation for the entire canopy, and increased soil water availability. We also expected that this eucalypt woodland, with its inherently low *L* [@Whitehead2004], would be even more responsive to eC~a~ than higher-*L* and more closed-canopy stands [@Norby2011]. Counter to our expectations but similar to Norby et al. [-@Norby2003], we found no significant effect of elevated C~a~ on *L*. It could be argued that the change in *L* in response to eC~a~ was too small to be detected. Natural variation in *L* among plots in a native woodland is indeed a challenge to manipulative studies with low replication, such as FACE experiments. To increase sensitivity and statistical robustness, we used a novel method to account for pre-treatment differences by using stand basal area as a covariate. Using this method, we estimated that we can detect a difference of only a `r format(Ldetect,digits=1)`% in *L*. Based on this small detectable difference, we conclude that the lack of a significant C~a~ effect is not due to a lack of power, but a true lack of response to elevated C~a~.

@@

\
Other studies in FACE are consistent with longer leaf-lifespan in elevated Ca shown here.
[@Taylor2008] reported delayed autumn senescence in *Populus* in two FACE experiments (PopFACE and AspenFACE). Although this indicates the canopy stayed greener longer in elevated C~a~, [@Gielen2001] reported faster turnover rate of leaves in elevated C~a~ due to overall higher productivity, leading to similar *L* in ambient C~a~ despite higher productivity.
But not with [@Warren2011]: increased senescence in eC~a~ in response to a combined drought and heat wave.

@@

**Figure 1**. (a) Target [CO~2~] (C~a~) relative to the ambient C~a~ treatment at the EucFACE. The elevated C~a~ treatment proceeded in steps of 30 ppm, reaching +150 ppm in Feb. 2013. (b) Canopy transmittance ($\tau_d$) estimated from continuous time series of photosynthetically active radition (PAR) below and above the canopy, for highly diffuse conditions (F~diff~ > 0.98) only. The solid lines are smoothed time series (with 95% confidence intervals) for each FACE ring (using a generalized additive model). Lines in blue colours denote ambient C~a~ treatment, red colours are elevated C~a~ (+150 ppm). Note steep increase in $\tau_d$ for Ring 6 around November 2014, due to an insect attack that resulted in near defoliation of the canopy. The arrow labelled ‘calibration’ indicates a drought period when $\tau_d$ steadily increased, used to calibrate *L* estimates to litter production (see Methods).

**Figure 2**. Average $L_\tau$ over the entire study period plotted against the basal area (sum of stem cross-sectional areas) of the FACE rings ($R^2$ = `r glance(balm)$r.squared`, *L* = `r tidy(balm)$estimate[1]` + `tidy(balm)$estimate[2]` BA). This relationship was used to calculate the *L* anomaly in Fig. 3, as the difference between *L* predicted from this regression and the actual $L_\tau$ at any given time.

@@

**Figure 6**.  (a) Total litter production (converted from data in Fig. 5 to an annual estimate) for each of the FACE rings plotted against mean $L_\tau$ during the entire litter collection period. The ratio of mean $L_\tau$ to annual litter production gives the mean leaf life span. Grey lines indicate leaf lifespan isolines. (b) Calculated leaf lifespan for the six FACE rings shows a significant difference between C~a~ treatments (p `r pval(LLtest$p.value)`).

**Figure 7**. Dynamics of *L* in relation to environmental drivers. (a) Smoothed time series of $L_\tau$ averaged across all six FACE rings. (b) From panel (a), we calculated the daily change in *L* (dL/dt), including the 95% confidence interval for the derivative of the time series. Green indicates significant positive dL/dt, red indicates significant negative dL/dt (at $\alpha$ = 0.05). Four of the five significant positive increases corresponded to approximate dates of commencement of leaf flushing (horizontal lines) observed in the field (except the third brief significant increase). (c) Soil water content estimated from over 0-80 cm soil depth interval, averaged across all sensors and all rings (blue line), and daily rainfall (black lines).  (d) Average daily air temperature with a fitted smoother.

@@

**Figure S2**. Correlation between *L* estimated from diffuse canopy transmittance ($L_\tau$) and from canopy photos (L~p~). Solid line is a 1:1 line, dashed line a regression line across all data (with the shading indicating the 95% confidence interval), and the colored lines regressions per plot across dates. (a) Estimates of L~p~ with the extinction coefficient (*k*) set to 0.5, as recommended by MacFarlane et al. [-@Macfarlane2007] ($L_\tau$ = `r tidy(lmcanphot)$estimate[1]` + `r tidy(lmcanphot)$estimate[2]` L~p~, R^2^ = `r glance(lmcanphot)$r.squared`, p `r pval(glance(lmcanphot)$p.value)`), (b) Estimates of L~p~ with *k* optimized to minimize the difference between L~p~ and $L_\tau$ ($L_\tau$ = `r tidy(lmcanphot2)$estimate[1]` + `r tidy(lmcanphot2)$estimate[2]`).

**Figure S3.** Comparison of canopy transmittance ($\tau$) during highly diffuse conditions (F~diff~ > 0.98), or using all PAR data without filtering. Shown are daily average $\tau$ for all six FACE rings (symbols), with a smoother (generalized additive model with 95% confidence intervals). Note the much lower $\tau$ during southern hemisphere winter when using all data compared to diffuse only. This difference arises due to low solar angles, which lead to higher attenuation of the direct beam as a result of a longer path length through the canopy.
